---
title: "Chapter 10: Join Revisited"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```

# Chapter 10: Joins revisited 

```{r inner_join_limit, echo=F}
#inner joins only return data for observations that exist in the joined data sets. For example, notice that 42 movies do not have any inventory information. 
dbListFields(conn = conn_mysql, name = 'inventory')
dbGetQuery(conn = conn_mysql, statement = "SELECT f.film_id, f.title, count(*) AS num_copies 
           FROM film AS f
           INNER JOIN inventory AS i
           ON f.film_id = i.film_id
           GROUP BY f.film_id, f.title")

#to return the number of copies for all movies, use the outer join. In this query, all the rows of the film data set are returned. Note that the calculation for num_copies
#has also been changed so that it is calculated by summing the number of times each inventory_id appears for each movie. If num_copies would be left as count(*), then null values would
#appear as 1 because they all appear in the film data set.
dbGetQuery(conn = conn_mysql, statement = "SELECT f.film_id, f.title,count(i.inventory_id) AS num_copies 
           FROM film AS f
           LEFT OUTER JOIN inventory AS i
           ON f.film_id = i.film_id
           GROUP BY f.film_id, f.title")

#the following two queries provide under-the-hood look for why num_copies returns as 0
dbGetQuery(conn = conn_mysql, statement = "SELECT f.film_id, f.title, i.inventory_id 
           FROM film AS f
           INNER JOIN inventory AS i
           ON f.film_id = i.film_id
           HAVING film_id 
           BETWEEN 13 AND 15")

dbGetQuery(conn = conn_mysql, statement = "SELECT f.film_id, f.title, i.inventory_id 
           FROM film AS f
           LEFT OUTER JOIN inventory AS i
           ON f.film_id = i.film_id
           HAVING film_id 
           BETWEEN 13 AND 15")

```

### Three-way outer joins

```{r three_outer_join, echo=T}
#the query below joins three data sets, with the rental data of each movie rental being extracted
dbListFields(conn = conn_mysql, name = 'rental')
dbGetQuery(conn = conn_mysql, statement = "SELECT COUNT(*) FROM film")
dbGetQuery(conn = conn_mysql, statement = "SELECT f.film_id, f.title, i.inventory_id, r.rental_date
           FROM film AS f
           LEFT OUTER JOIN inventory AS i
           ON f.film_id = i.film_id
           LEFT OUTER JOIN rental AS r
           ON i.inventory_id = r.inventory_id
           WHERE f.film_id BETWEEN 13 AND 15")
```

## Cross joins 

Cross joins return all possible combinations of the selected variables. 

```{r cross_join, echo=F}
dbGetQuery(conn = conn_mysql, statement = "SELECT c.name AS category_name, l.name AS language
           FROM category AS c
           CROSS JOIN language l")
```

Cross joins can be used to efficiently create tables with categories not directly calculated from the data. The query below implements a cross join to create a table where the number of
rentals on each day of 2020 is returned. The `DATE_ADD()` function is used to convert the numbers to day in the year 2020. 

```{r cross_join_ex, echo=F}
dbListFields(conn = conn_mysql, name = 'rental')
dbGetQuery(conn = conn_mysql, statement = "SELECT date(rental_date) FROM rental LIMIT 5")
dbGetQuery(conn = conn_mysql, statement = "SELECT days.date, COUNT(r.rental_id) AS num_rentals
           FROM rental AS r
           RIGHT OUTER JOIN
           (SELECT DATE_ADD('2005-01-01', INTERVAL(ones.num + tens.num + hundreds.num) day) AS date
           FROM
           (SELECT 0 num UNION ALL
           SELECT 1 num UNION ALL
           SELECT 2 num UNION ALL
           SELECT 3 num UNION ALL
           SELECT 4 num UNION ALL
           SELECT 5 num UNION ALL
           SELECT 6 num UNION ALL
           SELECT 7 num UNION ALL
           SELECT 8 num UNION ALL
           SELECT 9 num) ones
           CROSS JOIN
           (SELECT 0 num UNION ALL
           SELECT 10 num UNION ALL
           SELECT 20 num UNION ALL
           SELECT 30 num UNION ALL
           SELECT 40 num UNION ALL
           SELECT 50 num UNION ALL
           SELECT 60 num UNION ALL
           SELECT 70 num UNION ALL
           SELECT 80 num UNION ALL
           SELECT 90 num) tens
           CROSS JOIN
           (SELECT 0 num UNION ALL
           SELECT 100 num UNION ALL
           SELECT 200 num UNION ALL
           SELECT 300 num) hundreds
           HAVING date < '2006-01-01'
           ORDER BY date) AS days
           ON days.date = date(r.rental_date)
           GROUP BY days.date")
```

```{r chapter_10_exercises, echo=F}
#calculate the total amount of money spent by each customer 
dbListFields(conn = conn_mysql,name = 'payment')
dbGetQuery(conn = conn_mysql, statement = "SELECT c.customer_id, c.first_name, c.last_name, SUM(p.amount) AS total_spent
FROM customer AS c
LEFT OUTER JOIN payment AS p
ON c.customer_id = p.customer_id
GROUP BY c.customer_id")

#same query using right outer join 
dbGetQuery(conn = conn_mysql, statement = "SELECT c.customer_id, c.first_name, c.last_name, SUM(amount) AS total_spent
FROM customer AS c
RIGHT OUTER JOIN payment AS p
ON c.customer_id = .customer_id
GROUP BY c.customer_id")

#generate set that shows the number of customers in each spending bracket of 0-50, 50-100, 100-200, 200+
dbGetQuery(conn = conn_mysql,
           statement = "SELECT pymnt_grps.name, count(*) AS num_customers, 
           sum(pymnt.num_rentals)/count(*) AS avg_num_cust_transactions, 
           sum(pymnt.num_rentals) AS number_cat_rentals,
           avg(pymnt.tot_payments) AS avg_cust_revenue
           
           FROM (SELECT customer_id,
           count(*) AS num_rentals, sum(amount) AS tot_payments
           FROM payment
           GROUP BY customer_id) AS pymnt
           INNER JOIN 
           
           (SELECT 'Small Fry' AS name, 0 AS low_limit, 74.99 AS high_limit 
           UNION 
           SELECT 'Average Joes' AS name, 75 AS low_limit, 149.99 AS high_limit
           UNION 
           SELECT 'Heavy Hitters' AS name, 150 AS low_limit, 9999999.99 AS high_limit) AS pymnt_grps
           
           ON pymnt.tot_payments BETWEEN pymnt_grps.low_limit AND pymnt_grps.high_limit
           GROUP BY pymnt_grps.name")

```



