---
title: "Chapter 11: Conditional Logic"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```
# Chapter 11: Conditional logic

```{r searched_case_expression, echo=F}
#returns 'active' when active = 1
dbListFields(conn = conn_mysql, name = 'customer')
dbGetQuery(conn = conn_mysql, statement = "SELECT first_name, last_name,
           CASE 
            WHEN active = 1 THEN 'ACTIVE'   
            ELSE 'INACTIVE'
           END activity_type
           FROM customer")

#query returns the number of rentals for inactive customers using; note that you cannot exclude the active column 
dbGetQuery(conn = conn_mysql, statement = "SELECT first_name, last_name, num_rentals 
           FROM 
           (SELECT c.first_name, c.last_name, COUNT(*) AS num_rentals, c.active
           FROM customer AS c
           LEFT JOIN payment AS p
           ON c.customer_id = p.customer_id
           GROUP BY c.customer_id
           HAVING c.active = 0) AS num_rentals")

##we can use a case expression to return the number of rentals per customer for inactive customers
dbGetQuery(conn = conn_mysql, statement = "SELECT c.first_name, c.last_name, 
           CASE 
            WHEN active = 1 THEN 0
            ELSE
             (SELECT count(*) FROM rental AS r 
             WHERE r.customer_id = c.customer_id)
           END num_rentals
           FROM customer AS c
           HAVING num_rentals > 0")

```


## Examples of case expressions 
### Result set transformation

The query below return the number of rentals in each month (from May to July). Note that `monthname()` extracts only the month from the date.  

```{r example_case, echo=F}
dbGetQuery(conn = conn_mysql, "SELECT monthname(rental_date) AS rental_month, COUNT(*) AS num_rentals
           FROM rental 
           WHERE rental_date BETWEEN '2005-05-01' AND '2005-08-01'
           GROUp BY rental_month")

```

We can then use simple case expressions to convert this data from long to wide format.

```{r example_case_transform, echo=F}
dbGetQuery(conn = conn_mysql, statement = "SELECT 
           SUM(CASE WHEN monthname(rental_date) = 'May' THEN 1 ELSE 0 END) AS may_rentals,
           SUM(CASE WHEN monthname(rental_date) = 'June' THEN 1 ELSE 0 END) AS june_rentals,
           SUM(CASE WHEN monthname(rental_date) = 'July' THEN 1 ELSE 0 END) AS july_rentals
           FROM rental
           WHERE rental_date BETWEEN '2005-05-01' AND '2005-08-01'")
```

Case expressions can also be used to determine whether certain relations exist. The query below determines whether each actor has appeared in a movies with ratings of G, PG, and NC-17. The second query determines the availability of each film according to a categorization nomenclature.

```{r example_case_exist, echo=F}
dbGetQuery(conn = conn_mysql, statement = "SELECT a.first_name, a.last_name, 
                  CASE WHEN EXISTS (SELECT 1 FROM film_actor AS fa 
                  INNER JOIN film AS f
                  ON fa.film_id = f.film_id
                  WHERE fa.actor_id = a.actor_id AND 
                  f.rating = 'G') THEN 'Y' 
                  ELSE 'N'
                  END g_actor, 
                  
                  CASE WHEN EXISTS (SELECT 1 FROM film_actor AS fa
                  INNER JOIN film AS f 
                  ON fa.film_id = f.film_id
                  WHERE fa.actor_id = a.actor_id AND
                  f.rating = 'PG') THEN 'Y' ELSE 'N'
                  END pg_actor, 
                  
                  CASE WHEN EXISTS (SELECT 1 FROM film_actor AS fa
                  INNER JOIN film AS f 
                  ON fa.film_id = f.film_id
                  WHERE fa.actor_id = a.actor_id AND
                  f.rating = 'NC-17') THEN 'Y' ELSE 'N'
                  END nc_actor
                  
                  FROM actor AS a
                  WHERE a.last_name LIKE 'S%' OR a.first_name LIKE 'S%'
           ")

dbGetQuery(conn = conn_mysql, statement = "SELECT f.title, 
                  CASE  (SELECT count(*) FROM inventory AS i
                  WHERE i.film_id = f.film_id)
                  WHEN 0 THEN 'OUT OF STOCK'
                  WHEN 1 THEN 'SCARCE'
                  WHEN 2 THEN 'SCARCE'
                  WHEN 3 THEN 'AVAILABLE'
                  WHEN 4 THEN 'AVAILABLE'
                  ELSE 'COMMON' END film_availability
                  
                  FROM film AS f")
```

### Division-by-zero errors

Division by zero leads cells to be filled with `NULL` values. The query below calculates the average amount spent in each rental for each customer. Note that internal referencing cannot
be performed within qury. 

```{r division_zero_solution, echo=F}
dbListFields(conn = conn_mysql, name = 'payment')
dbGetQuery(conn = conn_mysql, statement = "SELECT c.customer_id, c.first_name, c.last_name, 
                  sum(p.amount) AS tot_payment,
                  count(p.amount) AS num_payments,
                  sum(p.amount) / 
                  CASE WHEN count(p.amount) = 0 THEN 1 ELSE count(p.amount)
                  END avg_payment
                  FROM customer AS c 
                  LEFT OUTER JOIN payment AS p 
                  ON c.customer_id = p.customer_id
                  GROUP BY c.customer_id")

```

### Conditional updates

The following query determines if customers are active by checking if they have rented a movie in the last 90 days.

```{r conditional_update, echo=F}
dbGetQuery(conn = conn_mysql, statement = "UPDATE customer AS c
           SET active = CASE 
           WHEN 90 <= (SELECT (datediff(now(), max(rental_date)))
           FROM rental AS r 
           WHERE r.customer_id = c.customer_id) THEN 0 ELSE 1 
           END 
           WHERE active = 1
           ")
```

## Exercises

```{r chapter-11-exercises, echo=F}
#assign encoding pattern to each actor
dbGetQuery(conn = conn_mysql, statement = "SELECT name, CASE name 
           WHEN  name in ('English', 'Italian', 'French', 'German') THEN 'latin1'
           WHEN name in ('Japanese', 'Mandarin') THEN 'utf8'
           ELSE 'unknown' 
           END 
           FROM language")

#convert the following query to wide format
dbGetQuery(conn = conn_mysql, statement = "SELECT 
           SUM(CASE WHEN rating = 'G' THEN 1 ELSE 0 END) AS num_g,
           SUM(CASE WHEN rating = 'PG' THEN 1 ELSE 0 END) AS num_pg,
           SUM(CASE WHEN rating = 'PG-13' THEN 1 ELSE 0 END) AS num_pg_13,
           SUM(CASE WHEN rating = 'R' THEN 1 ELSE 0 END) AS num_r,
           SUM(CASE WHEN rating = 'NC-17' THEN 1 ELSE 0 END) AS num_nc_17
           FROM film")
```

