---
title: "Chapter 13: Indexes and Constraints"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```
# Chapter 13: Indexes and constraints 

## Indexes 

Indexes are crucial for expediting SQL commands. For example, searching for names that begin with 'R' would take an overwhelming amount of time in a data set with 3 million rows. To 
expeite the searchtime, indexes can be used. Like an index in a book, an index allows an SQL command to only search a subset of rows (instead of every row). 

### Index creation

Indices can easily be created by nesting the `ADD INDEX` statement in an `ALTER TABLE` command. Having created a table for an index, the query optimizer can now determine whether queries
would benefit from using the index in searches. We can see that the `customer` table has indices for the `store_id`, `address_id`, and `last_name` columns. Note that, when creating 
tables, an index is automatically generated for the primary key column. 

```{sql index_creation, echo=F}
ALTER TABLE customer 
ADD INDEX idx_email (email); 
```

```{r view_index, echo=F}
dbListFields(conn = conn_mysql, name = 'customer')
dbGetQuery(conn = conn_mysql, statement = "SHOW INDEX FROM customer")
```

### Unique index

Some columns in datasets can be permitted to have identical values (e.g., first & last names), but other columns must have unique values (e.g., primary key, email). An index can be
designed to have unique values by simply adding the `UNIQUE` keyword in the index creation statements. 

```{r unique_index_creation, echo=F}
dbGetQuery(conn = conn_mysql, statement = "ALTER TABLE customer 
           ADD UNIQUE idx_email (email)")

#returns errors; Duplicate entry 'ALAN.KAHN@sakilacustomer.org' for key
dbGetQuery(conn = conn_mysql, statement = "INSERT INTO customer
          (store_id, first_name, last_name, email, address_id, active)
         VALUES
          (1,'ALAN','KAHN', 'ALAN.KAHN@sakilacustomer.org', 394, 1);
    ERROR 1062 (23000): Duplicate entry 'ALAN.KAHN@sakilacustomer.org'
      for key 'idx_email'")
```

### Multicolumn indexes

Multicolumn indexes can also be created. If you know queries will search for rows based on first and last names, the table can be equipped with an index comprised of the first and last name. Note that the order is important. If, for example, the rows are organized by last name, then `last_name` should preceed `first_name` in the index. 

```{sql multicolumn_index, echo=F}
ALTER TABLE customer
ADD INDEX idx_full_name (last_name, first_name);
```

### Types of indexes
#### B tree

B-tree index types are most common and constitute the default index type across all major database systems (Oracle, Microsoft SQL, MySQL). B-tree indices divide column values into 
alphabetical (or numerical) sets and then further subdivide these sets. Instead of searching through all values in a column, only the values in a leaf node are searched. Importantly, 
b-tree indices can withstand heavy modification by redistributing values across the branch and leaf nodes.

#### Bitmap indices

Indices are used for low-cardinality data. 

#### Text indices

Indexes are used for searching text content. 

### How indexes are used

Use the `EXPLAIN` keyword to see how SQL goes about using indexes. The execution path shown below indicates that the query used the `idx_last_name` key before the `full_name` key
was added, but that the `full_name` key was used after it was added. 

```{r index_checkup, echo=F}
dbGetQuery(conn = conn_mysql, statement = "EXPLAIN
SELECT customer_id, first_name, last_name
FROM customer
WHERE first_name LIKE 'S%' AND last_name LIKE 'P%'")

dbGetQuery(conn = conn_mysql, statement = "ALTER TABLE customer
ADD INDEX idx_full_name (last_name, first_name)")
```

### The downsides of indexes

Indexes can dramatically slow down the rate at which tables are modified because, with each modification, each key must be changed. Therefore, only add keys to tables when there is good
reason to do so. Otherwise, add and delete keys when they are only needed for temporary purposes. Two other rules to follow when using indexes are listed below:

1) Build indexes on all columns referenced by foreign keys. 
2) Index columns frequently used to retrieve data.

## Constraints

Three types of constraints:

1) Primary key constraints
2) Foreign key constraints: restrict one or more columns in a table to contain only values in another table's primary key columns (for example, changing a customer's ID should results
in changes to this ID value in all other tables)
3) Unique constraints restrict columns to have unique values. 
4) Check constraints restrict range of permissible values. 

### Constraint creation 

The SQL commands below place a foreign key on the `address_id` column of the `customer` table such that its values must be contained within the`address_id` column values of the `address`
table (parent table). Note that the InnoDb engine must be used to create foreign key constraints. The `on` clauses specify that deletions to `address_id` and `store_id` column values in the parent tables (`address` and `store`, respectively) will be blocked (`ON DELETE RESTRICT`) but that updates will be carried over (`ON UPDATE CASCADE`). 

```{sql constraint_creation, echo=F}
CREATE TABLE customer (
      customer_id SMALLINT UNSIGNED NOT NULL AUTO_INCREMENT,
      store_id TINYINT UNSIGNED NOT NULL,
      first_name VARCHAR(45) NOT NULL,
      last_name VARCHAR(45) NOT NULL,
      email VARCHAR(50) DEFAULT NULL,
      address_id SMALLINT UNSIGNED NOT NULL,
      active BOOLEAN NOT NULL DEFAULT TRUE,
      create_date DATETIME NOT NULL,
      last_update TIMESTAMP DEFAULT CURRENT_TIMESTAMP ON UPDATE CURRENT_TIMESTAMP,
      
      PRIMARY KEY (customer_id),
      KEY idx_fk_store_id (store_id),
      KEY idx_fk_address_id (address_id),
      KEY idx_last_name (last_name),
      CONSTRAINT fk_customer_address FOREIGN KEY (address_id)
      REFERENCES address (address_id) ON DELETE RESTRICT ON UPDATE CASCADE, 
      CONSTRAINT fk_customer_store FOREIGN KEY (store_id)
      REFERENCES store (store_id) ON DELETE RESTRICT ON UPDATE CASCADE)
ENGINE=InnoDB DEFAULT CHARSET=utf8;
```

The queries below provide two examples where deletions of `address_id` values in either the parent table (`customer`) or the child table (`address`) are blocked.

```{r view_foreign_keys, echo=F}
#we can see that the keys of the customer match how it was created above
dbGetQuery(conn = conn_mysql, statement = "SELECT constraint_name, table_name, constraint_type
        FROM information_schema.table_constraints
        WHERE table_schema = 'sakila' AND table_name = 'customer'
        ORDER BY 3,1")
```
Note that rows can be deleted from the child table (`customer` in this case) but not from the parent table (`address`). 

```{r blocked_deletions, echo=F}
dbListFields(conn = conn_mysql, name = 'customer')
dbGetQuery(conn = conn_mysql, statement = "USE sakila")

dbGetQuery(conn = conn_mysql, statement = "SELECT c.first_name, c.last_name, c.address_id, a.address
FROM customer c
INNER JOIN address a
ON c.address_id = a.address_id
WHERE c.address_id = 123")

#blocked deletion from parent table 
dbGetQuery(conn = conn_mysql, statement = "DELETE FROM address WHERE address_id = 123")

#permitted deletion from child 
dbGetQuery(conn = conn_mysql, statement = "DELETE FROM customer WHERE address_id = 123")
```
```{r permitted_updates, echo=F}
dbGetQuery(conn = conn_mysql, statement = "UPDATE address
           SET address_id = 9999 
           WHERE address_id = 123")
```

```{r chapter_13_exercises, echo=F}
#Generate an alter table statement for the rental table so that an error will be raised if a row having a value 
#found in the rental.customer_id column is deleted from the customer (parent) table.
dbGetQuery(conn = conn_mysql, statement = "ALTER TABLE rental
    ADD CONSTRAINT fk_customer_address FOREIGN KEY (customer_id)
    REFERENCES customer (customer_id) ON DELETE RESTRICT")

dbGetQuery(conn = conn_mysql, statement = "CREATE INDEX idx_payment01 
           ON payment (payment_date, amount)")

```


