---
title: "Chapter 14: Views"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```

# Chapter 14: Views

Views are indispensable for helping end users effectively interact with data. Views represent customized tables and, importantly, do not take up any memory in the hard disk because
they are simply reserved pieces of code. For example, the following query creates a view of the customer table with the email addresses anonymized. This anonymized view can be used
by all non-marketing personnel who do not need email information, keeping data security high. 

```{r view_example, echo=F}

dbGetQuery(conn = conn_mysql, statement = "CREATE VIEW customer_vw (
           customer_id, 
           first_name, 
           last_name, 
           email)
           AS SELECT  customer_id, first_name, last_name, 
           concat(substr(email,1,2), '*****', substr(email, -4)) email
           FROM customer")

#query view (this runs the code thatc reates the view and the commands that qierythe view)
dbGetQuery(conn = conn_mysql, statement = "SELECT first_name, last_name, email
           FROM customer_vw")

#use DESCRIBE to view the columns in a view (or dbListFields)
dbListFields(conn = conn_mysql, name = 'customer_vw')
dbGetQuery(conn = conn_mysql, statement = "DESCRIBE customer_vw")
```

Views can be used similarly to any other table. Select clauses can be applied to views and views can be joined to other tables. 

## Why use views?

### Data security

### Data aggregation

Views can be created so that end users can easily access pre-aggregated tables. If, at some future point, it would be more effective to put the view in a table, this can easily be accomplished. 

### Hiding complexity 

Some tasks require the linking of several tables (a very difficult and time-consuming task for end users). Complicated tables can be stored in views. For example, the following view 
stores the following information about each film: the film category, the number of actors appearing in the film, the total number of copies in inventory, and the number of rentals for
each film. 

```{r view_example_hiding_complexity, echo=F}
dbGetQuery(conn = conn_mysql, statement = " 
           CREATE VIEW film_report (
           AS 
            SELECT f.film_id, f.title, f.description, f.rating,
               #get film category for each film 
               (SELECT c.name 
               FROM category AS c 
                  INNER JOIN film_category AS fc
                  ON c.category_id = fc.category_id
               WHERE fc.film_id = f.film_id) AS category_name
            
            FROM film AS f")
```

Note that five variables are generated using subqueries and so, if these variables are not selected from the view, then the scalar subqueries will not be computed. 

### Joining partitioned data

## Updatable views

Views in MySQL and Oracle can be modified, but must follow the following rules:

1) No aggregate functions (max, min, avg).
2) The view does not contain `group by` or `having` clauses.
3) The view does not use `UNION`, `UNION ALL`, or `DISTINCT`.
4) The `from` clause includes at least one table or updatable view. 
5) The `from` clause uses only inner joins if there is more than one table or view. 
6) No subqueries exist in the select or from clause, and any subqueries in the where clause do not refer to tables in the from clause.

\noindent
Essentially, a view table can only be editable if it keeps the data at the lowest level. 

### Updating simple views

The query below updates a row in the `customer_vw` table. Note howthe update in the `customer_vw` view carries over into the customer table. Note that data cannot be inserted into views.

```{r update_simple_view, echo=F}
dbGetQuery(conn = conn_mysql, statement = "UPDATE customer_vw
           SET last_name = 'Smith-Allen' 
           WHERE customer_id = 1")

dbGetQuery(conn = conn_mysql, statement = "SELECT last_name FROM customer WHERE customer_id = 1")
```

### Updating complex views

```{r update_complex_view, echo=F}
dbGetQuery(conn = conn_mysql, statement = "CREATE VIEW customer_details
    AS
    SELECT c.customer_id,
      c.store_id,
      c.first_name,
      c.last_name,
      c.address_id,
      c.active,
      c.create_date,
      a.address,
      ct.city,
      cn.country,
      a.postal_code
    FROM customer c
      INNER JOIN address a
      ON c.address_id = a.address_id
      INNER JOIN city ct
      ON a.city_id = ct.city_id
    INNER JOIN country cn
      ON ct.country_id = cn.country_id;")

#data can only be updated and inserted in tables individually
dbGetQuery(conn = conn_mysql,statement = "SELECT * FROM customer_details LIMIT 10")
```

```{r echo=F, chapter_14_exercises}
#1) Create view that contains each film that every actor appears in (film_title, category_name, first_name, last_name)
#tables to use: actor --> film_actor --> film --> film_categoru 
dbListFields(conn = conn_mysql, name = 'actor')
dbGetQuery(conn = conn_mysql, statement = "SELECT title, category_name, first_name, last_name FROM actor_vw
WHERE last_name = 'FAWCETT'")

dbGetQuery(conn = conn_mysql, statement = "CREATE VIEW actor_vw 
                  AS
           SELECT f.title, a.first_name, a.last_name, c.name AS category_name
           FROM actor AS a
            INNER JOIN film_actor AS fa
            ON a.actor_id = fa.actor_id
            INNER JOIN film AS f 
            ON fa.film_id = f.film_id
            INNER JOIN film_category AS fc
            ON f.film_id = fc.film_id
            INNER JOIN category AS c
            ON fc.category_id = c.category_id
           ")

#2. The film rental company manager would like to have a report that includes the name of every country, along with the total payments for all customers who live in each country. Generate a view definition that queries the country table and uses a scalar subquery to calculate a value for a column named tot_payments.
#city --> address --> customer --> payment 

dbGetQuery(conn = conn_mysql, statement = "CREATE VIEW country_revenue_vw
           AS
           SELECT c.country, 
           (SELECT sum(p.amount) AS country_revenue
           FROM city AS ci 
               INNER JOIN address AS a 
               ON ci.city_id = a.city_id
               INNER JOIN customer AS cu
               ON a.address_id = cu.address_id
               INNER JOIN payment as p
               ON cu.customer_id = p.customer_id
           WHERE ci.country_id = c.country_id) AS tot_payments
           FROM country AS c")

#use 
dbGetQuery(conn = conn_mysql, statement = "
SELECT c.country, 
(SELECT sum(p.amount) AS country_revenue
           FROM city AS ci 
               INNER JOIN address AS a 
               ON ci.city_id = a.city_id
               INNER JOIN customer AS cu
               ON a.address_id = cu.address_id
               INNER JOIN payment as p
               ON cu.customer_id = p.customer_id
           GROUP BY ci.country_id) 
FROM country AS c")
```


