---
title: "Chapter 16: Analytic functions"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```

# Chapter 16: Analytic functions

Analytic functions are useful for summarizing data. The queries below show several examples of analytic functions. 

```{r analytic_function_examples, echo=F}
dbGetQuery(conn = conn_mysql, statement = 
           "SELECT quarter(payment_date) AS quarter, 
           monthname(payment_date) AS month, 
           sum(amount) AS monthly_sales
           FROM payment
           WHERE year(payment_date) = 2005
           GROUP BY quarter, month")

#this query gets the monthlyrevenue and also includes the max revenue in each quarter and over the four-month period
dbGetQuery(conn = conn_mysql, statement = 
           "SELECT quarter(payment_date) AS quarter, 
           monthname(payment_date) AS month, 
           sum(amount) AS monthly_sales, 
           max(sum(amount)) over() AS max_overall_sales, 
           max(sum(amount)) over(partition by quarter(payment_date)) AS max_qrtr_sales
           FROM payment
           WHERE year(payment_date) = 2005
           GROUP BY quarter, month")

#make a column for the rank order using rank()
dbGetQuery(conn = conn_mysql, statement = 
           "SELECT quarter(payment_date) AS quarter, 
           monthname(payment_date) AS month, 
           sum(amount) AS monthly_sales, 
           rank() over (order by sum(amount) desc) AS sales_rank
           FROM payment
           WHERE year(payment_date) = 2005
           GROUP BY quarter, month
           ORDER BY quarter, monthname(payment_date)
           ")

#create rank to be based on the quarter; SQL commands rank amounts of each month within each quarter 
dbGetQuery(conn = conn_mysql, statement = 
           "SELECT quarter(payment_date) AS quarter, 
           monthname(payment_date) AS month, 
           sum(amount) AS monthly_sales, 
           rank() over (partition by quarter(payment_date)
               order by sum(amount) desc) AS qrtr_rank
           FROM payment
           WHERE year(payment_date) = 2005
           GROUP BY quarter, month
           ORDER BY quarter, monthname(payment_date)
           ")

```

## Ranking 
### Ranking functions

* `row_number()` returns a unique number for each row. 
* `rank` returns the same ranking in case of a tie, with gaps in the rankings (next rank is the second-next integer value).
* `dense_rank` returns the same ranking in case of tie, with no gaps in the rankings (meaning the next rank is the next integer value). 

The query below counts the number of rentals for each customer, the customer's rank (in addition to the two types of ranks). 

```{r SQL_rank, echo=F}
#select top 10 customers based on number of rentals
dbGetQuery(conn = conn_mysql, statement = "
           SELECT customer_id, count(*) AS num_rentals, 
           row_number() over (order by count(*) desc) AS row_num_rank, 
           dense_rank() over (order by count(*) desc) AS dense_rnk
           FROM rental 
           GROUP BY customer_id
           LIMIT 10")

#rank customers based on total amount of money spent in each month. Use `partition by` to rank customers within each month. 
dbGetQuery(conn = conn_mysql, statement = " 
           SELECT c.first_name, c.last_name, monthname(payment_date) AS month,
           sum(p.amount) AS tot_pymt, 
           rank() over(partition by monthname(payment_date)
               order by sum(p.amount) desc) AS rank_rev
           FROM customer AS c 
            INNER JOIN payment AS p
            ON c.customer_id = p.customer_id 
           GROUP BY c.customer_id, monthname(payment_date) 
           ")

#to select the top 5customers of each month, we need to nest the ranking in a subquery  
dbGetQuery(conn = conn_mysql, statement = " 
           SELECT first_name, last_name, month, tot_pymt, rank_rev
           FROM 
           
               (SELECT c.first_name, c.last_name, monthname(payment_date) AS month, 
               sum(p.amount) AS tot_pymt, 
               rank() over(partition by monthname(payment_date)
               order by sum(p.amount) desc) AS rank_rev
               FROM customer AS c
                  INNER JOIN payment AS p
                  ON c.customer_id = p.customer_id
               GROUP BY p.customer_id,  monthname(payment_date)) cust_rank
          WHERE rank_rev <=5 
           ")

dbGetQuery(conn = conn_mysql, statement = "SELECT DISTINCT monthname(rental_date) FROM rental")
```

## Reporting functions

Aggregate functions can be used with `over` to calculate descriptive statistics for certain groups/ranges of time. The query below calculates the total revenue for each month.  

```{r reporting_functions, echo=F}
dbGetQuery(conn = conn_mysql, statement = "SELECT sum(amount) AS tot_revenue, monthname(payment_date) AS month
           FROM payment 
           GROUp BY month")

#return payment_month, monthly_total, & grand_total for payments above 10$, first_name, last_name (uses subquery)
dbGetQuery(conn = conn_mysql, statement = "SELECT first_name, last_name, amount, month, monthly_total, grand_total
           FROM 
               (SELECT c.first_name, c.last_name, p.amount,
               monthname(p.payment_date) AS month,
               sum(p.amount) over (partition by monthname(p.payment_date)) AS monthly_total, 
               sum(p.amount) over() AS grand_total
               FROM payment AS p
                  INNER JOIN customer AS c
                  ON p.customer_id = c.customer_id
               WHERE amount >= 10) AS monthly_info")

#does not use subquery 
dbGetQuery(conn = conn_mysql, statement = "SELECT c.first_name, c.last_name, p.amount,
               monthname(p.payment_date) AS month,
               sum(p.amount) over (partition by monthname(p.payment_date)) AS monthly_total, 
               sum(p.amount) over() AS grand_total
               FROM payment AS p
                  INNER JOIN customer AS c
                  ON p.customer_id = c.customer_id
               WHERE amount >= 10")

#return table with total monthly sales and percentage of total sales for each month
dbGetQuery(conn = conn_mysql, statement = "SELECT monthname(payment_date) AS month, 
           sum(amount) AS monthly_revenue,
           round(sum(amount)/sum(sum(amount)) *100, 2) AS pct_total
           FROM payment
           GROUP BY monthname(payment_date)")

#reporting with a descriptor column (month, monthly revenue, descriptor)
dbGetQuery(conn = conn_mysql, statement = "SELECT monthname(payment_date) AS month, 
           sum(amount) AS monthly_revenue,
           CASE sum(amount) 
               WHEN max(sum(amount)) over() THEN 'Highest'
               WHEN min(sum(amount)) over() THEN 'Lowest'
               ELSE 'Middle'
           END AS descriptor
           FROM payment 
           GROUP BY monthname(payment_date)")
```

### Window frames

Queries can also be equipped to calculate functions over desired periods of time. For example, the query below computes the rolling sum across each month. 

```{r window_frames, echo=F}
dbGetQuery(conn = conn_mysql, statement = "SELECT yearweek(payment_date) AS month, 
                  sum(amount) AS month_total, 
                  sum(sum(amount)) over (rows unbounded preceding) AS rolling_sum 
                  FROM payment
                  GROUP BY yearweek(payment_date)
           ")

#calculate rolling average over a three-week period 
dbGetQuery(conn = conn_mysql, statement = "SELECT yearweek(payment_date) AS week, 
           sum(amount) AS month_total, 
           avg(sum(amount)) over (rows between 1 preceding and 1 following) AS 3_wk_avg
           FROM payment
           GROUP BY week")

#calculate rolling weekly average. Note that range can be used to specify a date interval. Note that ORDER BY must be used with range. 
dbGetQuery(conn = conn_mysql, statement = "SELECT date(payment_date) AS date, 
           sum(amount), 
           avg(sum(amount)) over (order by date(payment_date) 
               range between interval 3 day preceding and interval 3 day following) AS week_rolling_avg
           FROM payment 
           GROUP BY date")
```

### Lag and lead

Queries can also compare values from previous rows. The query below calculates the revenue each week and the percentage difference from the week before. The `lag` keyword is used to 
calculate the revenue from the previous week and the `lead` keyword is used to calculate the revenue from the coming week. 

```{r lag_lead, echo=F}
#calculates weekly_revenue, revenue from previous and coming weeks
dbGetQuery(conn = conn_mysql, statement = "SELECT yearweek(payment_date) AS week, 
           sum(amount) AS weekly_revenue, 
           lag(sum(amount), 1) over(order by yearweek(payment_date)) AS prev_wk_total, 
           lead(sum(amount), 1) over(order by yearweek(payment_date)) AS next_wk_total
           FROM payment 
           GROUP BY yearweek(payment_date)")

#calculates weekly revenue and percentage change from previous week 
dbGetQuery(conn = conn_mysql, statement = "SELECT yearweek(payment_date) AS week, 
           sum(amount) AS weekly_revenue, 
           round(sum(amount)/lag(sum(amount), 1) over(order by yearweek(payment_date))*100, 2) AS pct_change
           FROM payment 
           GROUP BY yearweek(payment_date)
           ORDER BY 1")

#calculates weekly revenue and percentage difference from previous week 
dbGetQuery(conn = conn_mysql, statement = "SELECT yearweek(payment_date) AS week, 
           sum(amount) AS weekly_revenue, 
           round((sum(amount) - lag(sum(amount), 1) over (order by yearweek(payment_date)))/
               lag(sum(amount), 1) over(order by yearweek(payment_date))*100, 2) AS pct_difference
           FROM payment 
           GROUP BY yearweek(payment_date)
           ORDER BY 1")
```

### Column value concatenation

The `group_concat` function works a bit like `pivot_wider()` in that in converts a set of columns to a wide format. For example, the query below returns a the set of films with only
4 actors and the actor names. 

```{r column_value_concatenation, echo=F}
dbGetQuery(conn = conn_mysql, statement = "SELECT f.title, 
           group_concat(a.last_name, a.first_name order by a.last_name separator ',') AS actors
           FROM actor AS a 
            INNER JOIN film_actor AS fa
            ON a.actor_id = fa.actor_id
            INNER JOIN film AS f
            ON fa.film_id = f.film_id
          GROUP BY f.title
          HAVING count(*) = 4;")
```

## Exercises

```{sql chapter_16_exercises, echo=F}
--write a query that retrieves every row from Sales_Fact, and add a column to generate a ranking based on the tot_sales column values. The highest value should receive a ranking of 1, and the lowest a ranking of 24.

SELECT year_no, month_no, tot_sales, 
   rank() over(order by tot_sales desc)
   FROM Sales_Fact
   
--Modify the query from the previous exercise to generate two sets of rankings from 1 to 12, one for 2019 data and one for 2020.
SELECT year_no, month_no, tot_sales, 
   rank() over(partition by year_no order by tot_sales desc)
   FROM Sales_Fact
   
--Write a query that retrieves all 2020 data, and include a column that will contain the tot_sales value from the previous month.
SELECT year_no, month_no, tot_sales, 
   lag(tot_sales, 1) over(order by month_no) AS prev_month_sales
   FROM Sales_Fact
   WHERE year_no = '2020'
```

