---
title: "Chapter 2: Query Primer"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

# Chapter 2: Query primer 
## Query mechanics

When connecting to a databse, a connection ID is issued after a connection is made. This ID number will be useful in instances where errors are incurred and help is needed from a 
database administrator. This can be accessed using `dbGetInto()`. Like the primary key for any table in a relational databae management system, the connection ID of a MySQL server is 
generated in increasing order (beginning from 1). 

```{r connection_mechanics, echo=F}
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'sakila', password = 'tiramisu')
DBI::dbGetInfo(dbObj = conn_mysql)$threadId #connection ID
```

```{r query_info, echo=F}
query_df <- dbGetQuery(conn = conn_mysql, statement = "SELECT * FROM CATEGORY")
dbGetQuery(conn = conn_mysql, statement = "SELECT version(), user(), database()")
```

Some notes on the following queries: 

1) Using `DISTINCT` to check that no duplicates exist in a result set works, but is computationally taxing with larger data sets. 

```{r query_info, echo=F}
#create new columns and give them aliases (AS)
alias_df <- dbGetQuery(conn = conn_mysql, 
statement = "SELECT language_id,
'COMMON' language_usage, 
language_id * 3.14159 AS lang_pi_value
FROM language")

#remove duplication (DISTINCT)
distinct_df <- dbGetQuery(conn = conn_mysql, 
statement = "SELECT DISTINCT actor_id FROM film_actor ORDER BY actor_id")
```

In addition to accessing data from permanent tables (i.e., created with `CREATE`), the `FROM` clause can be used to access data from **derived tables** (i.e., tables in temporary memory),
**temporary tables**(tables created during the session that disappear after the session is closed), and **virtual tables** (i.e., created with `CREATE VIEW` [similar to using `View()` in
R]). The first query uses the `FROM` clause to access data from the `cust` derived table and then concatenates the first and last names from the derived table. The second query creates a
temporary table using `CREATE TEMPORARY`. The third query creates a virtual table that has no associated data and is stored in the data dictionary. 

```{r from_clause, echo=F}
#derived (sub-query) tables
derived_query_df <- dbGetQuery(conn = conn_mysql, statement = "SELECT CONCAT (cust.first_name, ',', cust.last_name) AS full_name FROM
(SELECT first_name, last_name, email FROM customer
WHERE first_name = 'JESSIE') cust")

#temporary table; note that by loading the temporary table into a dataframe in R, the data has been loaded into the temporary memory of the computer
dbExecute(conn = conn_mysql, 
statement = "CREATE TEMPORARY TABLE actors_j
(actor_id smallint(5),
first_name varchar(45),
last_name varchar(45))")

##insert rows from actor table into temporary table
dbExecute(conn = conn_mysql, 
statement = "INSERT INTO actors_j
(SELECT actor_id, first_name, last_name 
FROM actor
WHERE last_name LIKE 'J%')")

temporary_query_df <- dbGetQuery(conn = conn_mysql, statement = "SELECT * FROM actors_j")
rm(temporary_query_df) #ensures dataframe is removed from temporary memory of computer 

#virtual table; no result set emerges
dbExecute(conn = conn_mysql, statement = "CREATE VIEW cust_vw AS
SELECT customer_id, first_name, last_name, active
FROM customer")

#use FROM to view contents of cust_vw
virtual_query_df <- dbGetQuery(conn = conn_mysql, statement = "SELECT first_name, last_name FROM cust_vw WHERE active = 0")
rm(virtual_query_df)
```

### Table links 

When selecting columns from two or more tables, each column should be prefixed with the table name. The query below finds the customers who rented a DVD on June 14, 2005. The `INNER JOIN`clause is used to select rows with customer ID values in both tables. Because table names are repeated, it is efficient to alias table names. 

```{r table_link_ex, echo=F}
#table link; no table aliasing
df_no_aliasing <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT customer.first_name, customer.last_name, TIME(rental.rental_date) rental_time
           FROM customer
           INNER JOIN rental
           ON customer.customer_id = rental.customer_id
           WHERE date(rental.rental_date) = '2005-06-14'")

#table link; table aliasing
df_aliasing <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, TIME(r.rental_date) rental_time
FROM customer AS c
INNER JOIN rental AS r
ON c.customer_id = r.customer_id
WHERE date(r.rental_date) = '2005-06-14'")

all.equal(df_no_aliasing, df_aliasing) #TRUE
rm(df_no_aliasing, df_aliasing)


```
## The WHERE clause

## The GROUP BY and HAVING clauses

The `GROUP BY` clauses works just like the `dplyr::group_by()` and the `HAVING` clause works just like `dplyr::filter()`. The query below counts the number of rentals per customer and
only returns customers who have rented from than 40 films. 
```{r aggregating_data, echo=F}
aggregate_df <- dbGetQuery(conn = conn_mysql, 
                           statement = "SELECT c.first_name, c.last_name, count(*) AS num_films_rented
                           FROM customer AS c
                           INNER JOIN rental AS R 
                           ON c.customer_id = r.customer_id
                           GROUP BY c.first_name, c.last_name
                           HAVING num_films_rented >= 40")
rm(aggregate_df)

```


## The ORDER BY clause 

Note the nomenclature for ordering columns in opposite orders. 
```{r ordering_data, echo=F}
dbGetQuery(conn = conn_mysql, 
                           statement = "SELECT c.first_name, c.last_name, count(*) AS num_films_rented
                           FROM customer AS c
                           INNER JOIN rental AS R 
                           ON c.customer_id = r.customer_id
                           GROUP BY c.first_name, c.last_name
                           HAVING num_films_rented >= 40 
                           ORDER BY last_name DESC, num_films_rented ASC")
rm(ordered_df)
```

## Exercises

```{r exercise_queries, echo=F}
#1. Retrieve the actor ID, first name, and last name for all actors. Sort by last name and then by first name.
dbGetQuery(conn = conn_mysql, 
           statement = 'SELECT actor_id, first_name, last_name 
           FROM actor 
           ORDER BY last_name, first_name
           ')

#2. Retrieve the actor ID, first name, and last name for all actors whose last name equals 'WILLIAMS' or 'DAVIS'. 
##this can be achieved using WHERE or HAVING
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT actor_id, first_name, last_name 
           FROM actor 
           WHERE last_name = 'Williams' OR last_name = 'Davis'")

#3. Write a query against the rental table that returns the IDs of the customers who rented a film on July 5, 2005 (use the rental.rental_date column, and you can use the date() function to ignore the time component). Include a single row for each distinct customer ID.
dbListFields(conn = conn_mysql, name = 'rental')
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT DISTINCT(c.customer_id), c.first_name, c.last_name 
           FROM customer AS c
           INNER JOIN rental AS r
           ON c.customer_id = r.customer_id
           WHERE date(r.rental_date)  = '2005-07-05'")

#more efficient query if you just want to get the number of unique customers and not any identifying information 
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT DISTINCT(customer_id)
           FROM rental 
           WHERE date(rental_date)  = '2005-07-05'")
```
