---
title: "Chapter 4: Filtering"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

# Chapter 4: Filtering
## Filtering for rows that do not match a condition

Rows of tables can be filtered based on whether they do not match a condition using the following three methods:

1) `NOT` clause
2) `<>`
3) `!=`

```{r anti_match_filter, echo=F}
anti_match_not_df <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT first_name, last_name 
           FROM customer 
           WHERE NOT (first_name = 'Steven' AND last_name = 'Young')
           ORDER BY first_name ASC
           LIMIT 10")

anti_match_sign_df <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT first_name, last_name 
           FROM customer 
           WHERE first_name <> 'Steven' AND last_name <> 'Young'
           ORDER BY first_name ASC
           LIMIT 10")

anti_match_not_equal_df <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT first_name, last_name 
           FROM customer 
           WHERE first_name != 'Steven' AND last_name != 'Young'
           ORDER BY first_name ASC
           LIMIT 10")

all_equal(anti_match_not_df, anti_match_sign_df) #TRUE
```


## Delete rows using filter conditions

Note that committing these queries will permanently alter the data set and so the default auto-commit mode of the MySQL server should be changed so that these modifications can be 
undone. 

```{r delete_rows, echo=F}
delete_query <- 'DELETE FROM rental
WHERE year(rental_date) <> 2005 AND year(rental_date) <> 2006'
```

##The BETWEEN operator

Note that the lower limit should be specified first and the upper limit should be specified second. Also note that the `BETWEEN` operator includes values that match the the upper and
lower limits as valid. When working with dates, the inclusive nature of the `BETWEEN` operator is unintuitive. For example, to select films rented on either June 14 or June 15, the 
query must state `2005-06-14` as the lower limit and `2005-06-16` as the upper limit because, by default, dates begin on the first minute of the day (i.e., midnight). The `BETWEEN` operator can be used to specify ranges of dates, numbers, and strings. 

```{r between_operator, echo=F}
#date range
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT customer_id, rental_date 
           FROM rental 
           WHERE rental_date BETWEEN '2005-06-14' AND '2005-06-16'
           LIMIT 10")

#string range 
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT last_name, first_name 
           FROM customer 
           WHERE first_name BETWEEN 'Fa' AND 'Fr'
           LIMIT 10")

#numeric range
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT customer_id, payment_date, amount
           FROM payment 
           WHERE amount BETWEEN 10.0 AND 11.99
           LIMIT 10")
```

### The IN operator

The `IN` operator is an efficient way to match to multiple conditions and can also be used to filter subqueries and be used in conjunction with the `IN` operator. The 

```{r in_operator, echo=F}
#matching multiple conditions 
dbGetQuery(conn = conn_mysql, 
           statement =  "SELECT title, rating
           FROM film
           WHERE rating IN ('G','PG')
           LIMIT 10")

#filtering a subquery; 
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT title, rating 
           FROM film 
           WHERE rating IN (SELECT rating 
               FROM film
               WHERE title LIKE '%Pet')
           LIMIT 10")

#NOT IN query 
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT title, rating
           FROM film 
           WHERE rating NOT IN ('PG-13', 'R', 'NC-17')
           LIMIT 10")
```

## Matching conditions

Wildcard characters can be used in queries to filter for the several below components:

* Strings beginning/ending with a certain character or substring. 
* Strings containing character/substring. 
* Strings with a specific format.

\noindent
Regular expressions provide the most flexibility for filtering rows. 
```{r matching_conditions, echo=F}
#string with a specific format
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT last_name, first_name 
           FROM customer 
           WHERE last_name LIKE '_A_T%S'")

#regular expression that finds people with last names beginning with either Q or Y
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT last_name, first_name 
           FROM customer 
           WHERE last_name REGEXP '^[QY]'")

```

## NULL values

NULL values are used in instances when values are not applicable (e.g., employee ID value for a transaction that occurred at an ATM), values that are not yet known (e.g., return
date for a film), and when values are undefined (e.g., an account that is created for a product that has not yet been added to the database). NULL values behave differently from other 
values in the following ways:

1) An expression can never equal null; they can only *be* null (which is tested with `IS NULL`). 
2) Two nulls are never equal to each other. 

```{r null_queries, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT rental_id, customer_id
           FROM rental 
           WHERE return_date IS NULL
           LIMIT 10")

#find all rentals that were returned outside of May-August 2005 and that were not returned outside of this period
dbGetQuery(conn = conn_mysql,
           statement = "SELECT rental_id, customer_id
           FROM rental 
           WHERE return_date NOT BETWEEN '2005-05-01' AND '2005-09-01'
           OR return_date IS NULL
           LIMIT 10")
```

## Practice

```{r filtering_practice, echo=F}
#Q1
dbGetQuery(con = conn_mysql, 
statement = "SELECT payment_id, customer_id, amount, date(payment_date)
FROM payment 
WHERE payment_id BETWEEN 101 AND 120 AND
(amount > 8 OR date(payment_date) = '2005-08-23')")

#Q3: Construct a query that retrieves all rows from the payments table where the amount is either 1.98, 7.98, or 9.98.
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT payment_id, amount, date(payment_date)
           FROM payment
           WHERE amount IN (1.98, 7.98, 9.98)")

#Q4. Construct a query that finds all customers whose last name contains an A in the second position and a W anywhere after the A.
##Note that % includes any number of characters (including zero).
dbListTables(conn = conn_mysql)
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT customer_id, first_name, last_name 
           FROM customer 
           WHERE last_name LIKE '_A%W%'")
```

