---
title: "Chapter 5: Querying Multiple Tables"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```


# Chapter 5: Querying multiple tables 
## Inner joins

When querying data, information is often obtained from multiple tables using the `JOIN` and `ON` clauses. The query below obtains the names and addresses of all customers by finding
cases where rows in the `customer` and `address` tables have the same addresses. Note that, although the MySQL server computes inner joins by default, it is good practice to specify the
type of join. Although the below inner join can also be accomplished using the `WHERE` clause, this method is not portable across servers and loses clarity when more than two tables are
queried. Additionally, the second join method is not clear on the type of join being used. 

```{r, inner_join, echo=F}
multi_table_join_1 <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, a.address
           FROM customer AS c
           INNER JOIN address AS a 
           ON c.address_id = a.address_id")

multi_table_join_2 <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, a.address
           FROM customer AS c, address AS a
           WHERE c.address_id = a.address_id")

multi_table_join_3 <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, a.address
           FROM customer AS c 
           INNER JOIN address AS a
           ON c.address_id = a.address_id
           WHERE a.postal_code = 52137")

all_equal(multi_table_join_1, multi_table_join_2, multi_table_join_3)

dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, a.address
           FROM customer AS c
           INNER JOIN address AS a 
           ON c.address_id = a.address_id")
```

### Joining three of more tables

We need two specify two join types and use two `ON` subclauses when joining three tables. A three-table query is needed to obtain the city in which each customer resides. The `address_id`
foreign key column is first needed to traverse from the `customer` to the `address` table, at which point the `city_id` foreign key is used to traverse from the `address` table to the 
`city` table. Note that this query can be achieved using any of the following three queries. In each query, it is best to conceptualize the query as cutting the result set each time 
an inner join is computed. 

```{r three_join_query, echo=F}
#first: querying customer --> address --> city 
dbListFields(conn = conn_mysql, name = 'city')
straight_join_1 <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT cust.first_name, cust.last_name, a.address, c.city
           FROM customer AS cust
           INNER JOIN address AS a
           ON cust.address_id = a.address_id
           INNER JOIN city AS c
           ON a.city_id = c.city_id")

#second: querying city --> address --> customer
straight_join_2 <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT cust.first_name, cust.last_name, a.address, c.city
           FROM city AS c
           INNER JOIN address AS a
           ON a.city_id = c.city_id
           INNER JOIN customer AS cust
           ON cust.address_id = a.address_id")

#third: querying address --> city --> customer 
straight_join_3 <- dbGetQuery(conn = conn_mysql, 
           statement = "SELECT cust.first_name, cust.last_name, a.address, c.city
           FROM address AS a
           INNER JOIN city AS c
           ON a.city_id = c.city_id 
           INNER JOIN customer AS cust
           ON a.address_id = cust.address_id")

all_equal(straight_join_1, straight_join_2, straight_join_3) #TRUE 
```

\noindent
Each of the above queries return the same result set because SQL is a nonprocedural language: That is, SQL determines how to best execute the query. The MySQL server determines the 
driving table using statistics in the `sakila` database and then determines the order in which to determine the remaining tables. Note that the `STRAIGHT_JOIN` keyword can be used
if you want to ensure the joins follow a particular order (note that, in any of the above cases, forcing the join to follow by nominal order produces the same result set).

```{r straight_join, echo=F}
#customer --> address --> city 
dbGetQuery(conn = conn_mysql, 
           statement = " SELECT STRAIGHT_JOIN cust.first_name, cust.last_name, c.city 
           FROM customer AS cust
           INNER JOIN address AS a
           ON cust.address_id = a.address_id
           INNER JOIN city AS c
           ON a.city_id = c.city_id")

#querying city --> address --> customer
dbGetQuery(conn = conn_mysql, 
           statement = " SELECT STRAIGHT_JOIN cust.first_name, cust.last_name, c.city 
           FROM city AS c
           INNER JOIN address AS a
           ON c.city_id = a.city_id
           INNER JOIN customer AS cust
           ON a.address_id = c.address_id")

#third: querying address --> city --> customer 




```

### Using the same table twice

The query below finds all films in which either the actors Cate McQueen or Cuba Birch appeared. The result set will include the actor's name (`first_name`, `last_name`) and the film name
(`title`). The second query finds the films in which both actors appeared. 

```{r query_same_table, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT a.first_name, a.last_name, f.title
           FROM film AS f
           INNER JOIN film_actor AS fa
           ON f.film_id = fa.film_id
           INNER JOIN actor AS a
           ON fa.actor_id = a.actor_id
           WHERE (a.first_name = 'Cate' AND a.last_name = 'McQueen') OR 
           (a.first_name = 'Cuba' AND a.last_name = 'Birch')
           LIMIT 10")

#find films in which both actors appear 
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT f.title, a1.first_name, a1.last_name, a2.first_name, a2.last_name
           FROM film AS f
           INNER JOIN film_actor AS fa1
           ON f.film_id = fa1.film_id
           INNER JOIN actor AS a1
           ON fa1.actor_id = a1.actor_id
           INNER JOIN film_actor AS fa2
           ON f.film_id = fa2.film_id
           INNER JOIN actor AS a2
           ON fa2.actor_id = a2.actor_id
           WHERE (a1.first_name = 'CATE' AND a1.last_name = 'MCQUEEN')
           AND (a2.first_name = 'CUBA' AND a2.last_name = 'BIRCH')")
```

## Self joins

```{r self_join, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "DESC film")

```

## Exercises
```{r querying_tables_questions, echo=F}
#2. Write a query that returns the title of every film in which an actor with the first name JOHN appeared.
dbListFields(conn = conn_mysql, name = 'city')
dbListTables(conn = conn_mysql)
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT f.title, a.first_name, a.last_name
           FROM film AS f
           INNER JOIN film_actor AS fa
           ON f.film_id = fa.film_id
           INNER JOIN actor AS a
           ON fa.actor_id = a.actor_id
           WHERE a.first_name = 'John'")

#3. Construct a query that returns all addresses that are in the same city. You will need to join the address table to itself, and each row should include two different addresses.
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT a1.address, a2.address, c.city
           FROM address AS a1
           INNER JOIN address AS a2
           ON a1.city_id = a2.city_id AND
           a1.address_id != a2.address_id
           INNER JOIN city AS c
           ON a1.city_id = c.city_id
           ")

#4. Find films in which both actors Cate McQueen and Cuba Birch appear 
dbListFields(conn = conn_mysql, name = 'film_actor')
#first, we find all rows in the film table that have at least two rows in the film_actor table
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT f.title, a1.first_name, a1.last_name, a2.first_name, a2.last_name
           FROM film AS f
           INNER JOIN film_actor AS fa1
           ON f.film_id = fa1.film_id  
           INNER JOIN actor AS a1
           ON fa1.actor_id = a1.actor_id
           
           INNER JOIN film_actor AS fa2
           ON f.film_id = fa2.film_id
           INNER JOIN actor AS a2
           ON fa2.actor_id = a2.actor_id #produces all possible two-actor combinations for each film 
           WHERE (a1.first_name = 'CATE' AND a1.last_name = 'MCQUEEN')
           AND (a2.first_name = 'CUBA' AND a2.last_name = 'BIRCH')")
```


