---
title: "Chapter 6: Working With Sets"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```
# Chapter 6: Working with sets

To perform set operations across two or more data sets, each data set must have the same number of rows and the data types of each column across the two data sets must be the same (or 
the server must be able to convert one to the other). 

## Set operators
### The UNION operator

Use `UNION` operator removes duplicates and the `UNION ALL` operator does not. The query below selects all the customer and actor names and creates a new column to track the type of 
name (i.e., `CUST` OR `ACTR`). The second query performs the same computation, but only finds instances where the first and last names begin with 'J' and 'D', respectively. Note that,
if the column delineating the type of observation was included in the second query, then there would be 5 rows instad of four because Jennifer Davis is a name contained in the `actor` and
`customer` tables.

```{r union_operator, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT 'CUST' typ, c.first_name, c.last_name 
           FROM customer AS c
           UNION
           SELECT 'ACTR' typ, a.first_name, a.last_name 
           FROM actor AS a")

dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name 
           FROM customer AS c
           WHERE c.first_name LIKE 'J%' AND c.last_name LIKE 'D%'
           UNION
           SELECT a.first_name, a.last_name 
           FROM actor AS a
           WHERE a.first_name LIKE 'J%' AND a.last_name LIKE 'D%'")
```

### The INTERSECT operator (INNER JOIN in MySQL)

The `INTERSECT` operator has been discontinued and so more clever ways of computing it must be devised. The following query finds actor and customer names that exist in both tables and
where the first name begins with 'J' and the last name begins with 'D'. 

```{r pseudo_intersect, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT a.first_name, a.last_name
           FROM actor AS a
           INNER JOIN customer AS c
           ON a.first_name = c.first_name AND a.last_name = c.last_name
           WHERE a.first_name LIKE 'J%' AND a.last_name LIKE 'D%' 
           ")
```

### The EXCEPT operator (LEFT JOIN in MySQL)

Like the `INTERSECT` operator, MySQL does not use the `EXCEPT` operator. In Oracle, the minus sign must be used. Nonetheless, workarounds can be generated to compute the `EXCEPT`
operator, which returns observations unique to a result set. The below query finds names in the `actor` and `customer` tables where the first name begins with 'J' and the last name begins with 'D' that that are unique to the `actor` table.  

```{r pseudo_intersect, echo=F}
#query does not eliminate names in both tables
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT a.first_name, a.last_name
           FROM actor AS a
           LEFT JOIN customer AS c 
           ON a.first_name = c.first_name and a.last_name = c.last_name
           WHERE (a.first_name LIKE 'J%' AND a.last_name LIKE 'D%')")

```

## Set operation rules

You can only order by column names in the first query. Therefore, give the same alias names to column names in each query. 

```{r sorting_compound_queries, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT a.first_name AS fname, a.last_name AS lname
           FROM actor AS a
           WHERE a.first_name LIKE 'J%' AND a.last_name LIKE 'D%'
           UNION
           SELECT c.first_name AS fname, c.last_name AS lname
           FROM customer AS c
           WHERE c.first_name LIKE 'J%' AND c.last_name LIKE 'D%'
           ORDER BY lname, fname")

```

## Set operation precedence


