---
title: "Chapter 8: Grouping and Aggregates"
header-includes:
   - \usepackage{nccmath}
   - \usepackage{caption}
   - \usepackage{longtable}
   - \usepackage{setspace}
   - \usepackage{pdflscape}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage{amsthm}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
   - \usepackage{upgreek}  #required for non-italicized Greek letters
   - \usepackage{subcaption}
   
      #- \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{labelfont={normalfont, bf}, singlelinecheck=false, labelsep=newline}
   
   - \DeclareCaptionFont{figCaptionFont}{\fontfamily{phv}} #sets caption font to sans serif font of Helvetica 
   - \DeclareCaptionFont{figCaptionSize}{\footnotesize} #set caption font size to footnote 
   - \DeclareCaptionFont{figCaptionStyle}{\textup}  #set caption font to non-italicized font  
   - \DeclareCaptionLabelSeparator{captionSep}{\newline\newline} #separates figure label and figure title with required white space
   - \captionsetup[figure]{font = {figCaptionFont,figCaptionSize, figCaptionStyle}, labelsep = captionSep,  justification=justified}

#setup for tables
   - \captionsetup[table]{font = {figCaptionFont,figCaptionSize,figCaptionStyle}, labelfont={bf}, labelsep=captionSep, justification = raggedright, margin = {0cm,0cm}}
   - \setlength\parindent{24pt}
   
format: "pandoc"
output: pdf_document
---

```{r setup, include=F}
knitr::opts_knit$set(root.dir = '~/Desktop/Datacamp/SQL')

library(easypackages)
packages_to_load <- c('tidyverse', 'RSQLite', 'dbplyr', 'DBI', 'odbc', 'RMySQL', 'reticulate', 'kableExtra', 'knitr', 'magick')
packages(packages_to_load)

reticulate::use_python(python = "/usr/local/msql/bin/")
knitr::knit_engines$set(python.reticulate = FALSE)
```

```{r server_connection, echo=F}
#set up connection with MySQL server; connect to meta data database
conn_mysql <- dbConnect(drv = MySQL(), user = 'root', port = 3306, dbname = 'information_schema', password = 'tiramisu')

dbGetQuery(conn = conn_mysql, statement = "USE sakila")
#set up connection with SQLite server
conn_sqlite <- dbConnect(RSQLite::SQLite(), 'professor_data/unitransparenz.sqlite')
```
# Chapter 8: Grouping and aggregates

## Grouping concepts

Although data are generally stored at their lowest level, we often need to analyze data at higher levels. To group data, use the `GROUP BY` clause. The query below counts the number of 
films rented by each customer. When dealing with groups, note that the `HAVING` keyword must be used instead of the `WHERE` keyword because `WHERE` must be placed before `GROUP BY`. 

```{r grouping_data, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, count(*) AS num_films_rented
           FROM rental AS r
           INNER JOIN customer AS c
           ON r.customer_id = c.customer_id
           GROUP BY r.customer_id
           HAVING num_films_rented > 40
           ORDER BY num_films_rented DESC
           ")

```

## Aggregate functions

The query below computes the following values for each customer: 

1) Minimum and maximum payment amounts of each customer (i.e., many instances where customers rent several films at once).
2) Average payment amount of each customer.
3) Total (sum) amount of money spent by each customer.
4) The number of films rented. 

```{r aggregate_functions, echo=F}
dbListFields(conn = conn_mysql, name = 'payment')

dbGetQuery(conn = conn_mysql, 
           statement = "SELECT c.first_name, c.last_name, 
           avg(amount) AS avg_payment_amount, 
           max(amount) AS max_payment_amount,
           min(amount) AS min_payment_amount, 
           sum(amount) AS total_money_spent,
           count(*) AS num_films_rented
           FROM rental AS r
           INNER JOIN customer AS c
           ON r.customer_id = c.customer_id
           INNER JOIN payment AS p
           ON c.customer_id = p.customer_id
           GROUP BY r.customer_id
           HAVING num_films_rented > 40
           ORDER BY num_films_rented DESC")
```

### Counting distinct values

Using `DISTINCT` with `COUNT` is used in the below query to get the sum of rentals for each customer, the number of times each rental value is observed, and the average amount of money
spent in each rental group. In interpreting this result set, we can see that 54 customers rented 26 movies and contributed $5840.96 in revenues. 

```{r distinct_count, echo=F}
dbListFields(conn = conn_mysql, name = 'payment')
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT cust.num_rentals, count(*) AS rental_group_size, sum(cust.amount) AS rental_group_revenue 
           FROM (SELECT count(*) AS num_rentals, sum(amount) AS amount
           FROM payment
           GROUP BY customer_id) AS cust
           GROUP BY cust.num_rentals
           ORDER BY rental_group_size DESC")
```

### Using expressions 

Expressions can be used when aggregating to extract more nuanced information from a data set. For example, the query below produces a result set that computes the average return time
for each customer and then computes the frequency with which each value is observed across all customers. 

```{r using_expressions, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT cust.avg_return_time, count(*) AS return_time_freq FROM 
           (SELECT round(avg(datediff(return_date, rental_date)), 1) AS avg_return_time
           FROM rental AS r
           GROUP BY r.customer_id) AS cust
           GROUP BY cust.avg_return_time")

```

### How nulls are handled 

Like telling arithmetic functions in R to ignore NA values. The below query counts the number of NULL values in the `payment_date` column. 

```{r counting_nulls, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT count(*) FROM payment
           WHERE payment_date IS NULL")

```

## Generating groups
#### Grouping via expressions

The query below groups via the year of the rental date. 

```{r group_expression, echo=F}
dbGetQuery(conn = conn_mysql,
           statement = "SELECT extract(YEAR FROM rental_date) AS year, COUNT(*) AS num_rentals
           FROM rental
           GROUP BY year")
```
### Generating rollups

Use the `rollup` keyword to get the total number of counts across all groups specified by the `GROUP BY` clause. The query below counts the number of films each actor appears in and also returns the total number of films (i.e., the sum of all the values in the `num_films` column). 

```{r rollups, echo=F}
dbGetQuery(conn = conn_mysql, 
           statement = "SELECT fa.actor_id, count(*) AS num_films
           FROM film_actor AS fa
           INNER JOIN film AS f
           ON fa.film_id = f.film_id
           GROUP BY fa.actor_id")
```


## Exercises 

```{r chapter_8_exercises, echo=F}
#Construct a query that counts the number of rows in the payment table.
dbGetQuery(conn = conn_mysql,
           statement = "SELECT COUNT(*) AS num_rows FROM payment")

#Modify your query from Exercise 8-1 to count the number of payments made by each customer. Show the customer ID and the total amount paid for each customer.
dbGetQuery(conn = conn_mysql,
           statement =  "SELECT COUNT(*) AS num_payments 
           FROM payment 
           GROUP BY customer_id")

#Modify your query from Exercise 8-2 to include only those customers who have made at least 40 payments.
dbGetQuery(conn = conn_mysql,
           statement =  "SELECT COUNT(*) AS num_payments 
           FROM payment 
           GROUP BY customer_id
           HAVING num_payments >= 40
           ORDER BY num_payments")
```


